---
title: "Bayesian Improved Surname Geocoding (BISG)"
author: 
date: "8/5/2020"
output: html_document
---

## What is Bayesian Improved Surname Geocoding?

Bayesian Improved Surname Geocoding (BISG) is a method that applies the Bayes Rule/Theorem to predict the race/ethnicity of an individual using the individual's surname and geocoded location [cite Elliott et. al 2008, Elliot et al. 2009, Imai and Khanna 2016]. Specifically, BISG identifies the prior probability or likelihood of a an individual's race given the racial/ethnic proportions present in the geographic location in which that person lives. The prior probability is then updated using the probability of an innvididual being of a certian race given their surname. These probabilities are applied to Bayes Rule using the following notation:

      [insert formula and annotations for BISG]

In R, the package that performs BISG is called, WRU: Who Are You `wru` [cite WRU package]. This vignette will walk you through how to prepare your geocoded voter file for performing BISG by stepping you throught the processing of cleaning your voter file, prepping voter data for running the BISG, and finally, performing BISG to obtain racial/ethnic probailities of individuals in a voter file.

## How to perform BISG on your voter file data?
The first step in performing BISG is to geocode your voter file addresses. For information on geocoding, visit the Geocoding Vignette. 

In this tutorial, we will be using the East Ramapo, New York voter file called, **voter_file_geocoded**, that consists of indivduals registering to vote during the year of 2016. 

Let's begin by loading your geocoded voter data into R/RStudio.

### Step 1: Load R libraries/packages, voter file, and census data
```{r}
# Load libraries/packages
library(devtools)
# install.packages("tidyverse")
library(tidyverse)
# install.packages("stringr")
library(stringr)
# install.packages("sf")
library(sf)
library(eiCompare)
library(wru)
library(readr)
```

Load in East Ramapo, NY School District voter registation data.
```{r}
# Load geocoded voter registration file
read_csv("~/shared/east_ramapo/data/voter_file_geocoded.csv")
```

Make sure to load your census data that details certain geographies (i.e. counties, cities, tracts, blocks, etc.)
```{r}
# Load New York census data
# census_data <- load("ny_census2.RData")
```

Next, load the state shape file using the sf::st_read() function.
```{r}
# Load New York block shape file
ny_shape <- st_read("~/shared/east_ramapo/data/tl_2014_36_tabblock10.shp")
```

Obtain the first six rows of the voter file to check that the file has downloaded properly.
```{r}
# Check the first 6 rows of the voter file
head(voter_file_geocoded)
```

Check the dimensions (the number of rows and columns) of the voter file.
```{r}
# Get the dimensions of the voter file
dim(voter_file_geocoded)
```
There are 13005 voters (or observations) and 14 columns in the voter file.

View the column names of the voter file. Some of these columns will be used along the journey to performing BISG.
```{r}
# Find out names of columns in voter file
names(voter_file_geocoded)
```

### Step 2: De-duplicate the voter file.

The next step involves removing duplicate voter IDs from the voter file, using the `dedupe_voter_file` function. Check the column name for the unique identifier assigned to each voter the voter file. For the East Ramapo voter file we are using, voter_id is the column named `id`.

```{r}
# Remove duplicate voter IDs (the unique identifier for each voter)
voter_file_geocoded <- dedupe_voter_file(voter_file = voter_file_geocoded, voter_id = "id")

# Check new dimensions of voter file after removing duplicate voter IDs
dim(voter_file_geocoded)
```
There are no duplicate voter IDs in the dataset for the East Ramapo school district data set.

### Step 3: Merge voter file and shape files. This function may take a minute to complete.
```{r}
# Load shape file
voter_shape_merged <- merge_voter_file_to_shape(
  voter_file = voter_file_geocoded,
  shape_file = ny_shape,
  crs = "+proj=longlat +ellps=GRS80",
  coords = c("lon", "lat"),
  voter_id = "id"
)

# Check first 6 rows of merged voter file.
head(voter_shape_merged)
```
```{r}
# Obtain dimensions of merged file.
dim(voter_shape_merged)
```

```{r}
# Get column names of merged file.
names(voter_shape_merged)
```


### Step 4: Extracting necessary columns from the voter file to perform BISG
The voter file contains lots of information about the voter. However, all information is not needed and some columns like voter ID and state.

```{r}
voter_file_to_bisg <- tidy_voter_file_wru(
  voter_file = voter_shape_merged,
  voter_id = "id",
  surname = "last",
  state = "STATEFP10",
  county = "COUNTYFP10",
  tract = "TRACTCE10",
  block = "BLOCKCE10"
)

# Check the first 6 rows of the voter_file_to_bisg
head(voter_file_to_bisg)
```
```{r}
dim(voter_file_to_bisg)

# Get column names for the voter file that is now prepped to perform BISG.
names(voter_file_to_bisg)
```

### Step 5: Perform BISG and obtain the predicted race/ethnicity of each voter.
```{r}
# Convert the voter_shaped_merged file into a data frame for performing BISG.
voter_file_complete <- as.data.frame(voter_shape_merged)
class(voter_file_complete)
```

```{r}
# Perform BISG
bisg_file <- eiCompare::wru_predict_race_wrapper(
  voter_file = voter_file_complete,
  census_data = census_data,
  voter_id = "id",
  surname = "last",
  state = "NY",
  county = "COUNTYFP10",
  tract = "TRACTCE10",
  block = "BLOCKCE10",
  census_geo = "block",
  use_surname = TRUE,
  surname_only = FALSE,
  surname_year = 2010,
  use_age = FALSE,
  use_sex = FALSE,
  return_surname_flag = FALSE,
  return_geocode_flag = FALSE,
  verbose = FALSE
)
head(bisg_file)
```
The BISG voter data.frame is now split into two tables `bisg_file$voter_file` and `bisg$bisg`. The racial proportions for every indidivual given their last name and location is generated and can now be used to perform ecological inference using eiCO=ompare.

## Summarizing and Plotting BISG output

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
