---
title: "eiCompare: Ecological Inference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ei}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette illustrates how we can use the Ecological Inference (EI) tools 
within `eiCompare` to learn about the voting behavior of different groups of 
voters. 

## The ecological inference problem

Ecological Inference involves estimating the voting behaviors of different 
groups of voters using only data about election results and the number of people
from each group that turned out to vote. Elections in the United States, and 
much of the rest of the world, use secret ballots that ensure confidentiality 
for every voter. Voter confidentiality is important for a functioning democracy, 
but it presents a problem for those who want to learn about voter behavior. We cannot directly 
measure how people vote in an election. Intead, we have to make inferences about how they might have voted 
using election results and information about voter turnout. We often end up with
datasets like the following, which contains aggregate election results and turnout 
information for each election precinct in Gwinnett County, Georgia:

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options(scipen = 999, digits = 4)
```

```{r}
suppressPackageStartupMessages({
  library(eiCompare)
  library(dplyr)
  library(doParallel)
})

data("gwinnett")
head(gwinnett)
```

This dataset contains the following columns:

- `precinct` - This uniquely identifies each precinct in the county.
- `turnout` - The total number of voters who cast ballots on election day
- `kemp`, `abrams`, `metz` - These count the number of votes cast for each
candidate.
- `white`, `black`, `hispanic`, `other` - These count the number of voters who
self-reported as each race group.

Election results by precinct are generally accessible online from election 
administrators. Measuring the turnout of different racial/ethnic groups, 
however, can present more of a challenge. Luckily, the state of Georgia requires
voters to self-report their race when they register. This means anyone can 
tabulate the number of voters self-reporting with each group in each precinct, 
provided they gain access to a voter file. That's exactly what we've done here.
To learn more about how to estimate voters' race based only on their name and 
address, see the tutorials on [Geocoding voter files](geo.com) and 
[Bayesian Improved Surname Geocoding](BISG.com)

We know the number of votes cast for each candidate and by each racial/ethnic 
group. From this data, we want to learn whether voters from different ethnic
groups prefer different candidates. 

## Preparing the data

First, we need to clean and prepare the data. `eiCompare` contains functions to
streamline data preparation and prevent common errors that emerge during this 
process.

First, let's set up our column vectors. Most `eiCompare` functions for ecological inference require these as parameters.

```{r}
cands <- c("kemp", "abrams", "metz")
races <- c("white", "black", "hispanic", "other")
total <- "turnout"
id <- "precinct"
```

We start by dealing with all the missing values in the our dataset. For this, we can use the function `resolve_missing_vals()` from the `eiCompare` package.

```{r}
gwinnett <- resolve_missing_vals(
  data = gwinnett,
  cand_cols = cands,
  race_cols = races,
  totals_col = total,
  na_action = "DROP"
)
```
Looks like this dataset doesn't contain any missing values. We can proceed to the next step. Next, we use `dedupe_precincts()` to check that the data does not contain any 
duplicate precincts.

```{r}
gwinnett <- dedupe_precincts(
  data = gwinnett,
  id_cols = id
)
```
We got a warning indicating that some rows of the data do appear to belong to the same precinct. When `dedupe_precincts()` identifies duplicates that it does not know how to resolve, the function returns a column called `duplicate` that highlights the duplicated rows. This helps us investigate the duplicates and try to figure out where they came from. Let's take a look at these now.

```{r}
gwinnett %>%
  filter(duplicate)
```

Here we see that precinct `035` appears to have entered the data twice. This anomaly has at least two possible explanations:

1. For most elections, precincts report their own results and data entry is often done by hand. This can lead to small mistakes like a precincts' vote tallies being split over two rows.

2. Election results databases typically contain the resutls of many elections that took place in different jurisdictions and at different times. When extracting the results of a particular election, an easy mistake is to accidentally extract some results from other elections, leading to repeated precinct IDs. 

In this case, we have confidence that this duplicate has emerged through an error in precinct reporting. We can combine the two columns using this code:

```{r}
missing_inds <- which(gwinnett$duplicate)
columns_to_add <- c(total, cands)
gwinnett[missing_inds[1], columns_to_add] <-
  gwinnett[missing_inds[1], columns_to_add] +
  gwinnett[missing_inds[2], columns_to_add]
gwinnett <- gwinnett[-missing_inds[2], ]
gwinnett[missing_inds[1], ]
```
Now we see the vote totals from the previous two columns have been summed together in this new row. We can double check that this eliminated all duplicates be running `dedupe_precints()` again.

```{r}
gwinnett <- dedupe_precincts(
  data = gwinnett,
  id_cols = id
)
```
Now we're sure that our dataset does not contain any missing values or duplicated precincts. We can proceed with standardizing the data. To execute ecological inference cleanly, we need to make sure to represent our data in proportions that sum to one. Not doing this can cause the EI functions to break or behave unexpectedly. For this, we can use `stdize_votes_all()`. 


```{r}
gwinnett_ei <- stdize_votes_all(
  data = gwinnett,
  cand_cols = cands,
  race_cols = races,
  totals_col = total
)
head(gwinnett_ei)
```


This function has produced a new dataframe where each column is a proportion. We can check that for each row, the candidate columns sum to one and the race columns sum to one using the following code.

```{r}
cand_sums <- sum_over_cols(data = gwinnett_ei, cols = cands)
race_sums <- sum_over_cols(data = gwinnett_ei, cols = races)
table(cand_sums, race_sums)
```
Here we see that, for each row in the dataset, the sum of the race columns equals one, and the sum of the candidate columns also equals one. We *always* need to satisfy this condition before running the EI functions.

We now have a new dataframe without missing values, with no duplicated precincts, and with standardized candidate and racial vote proportion colums. We can begin conducting ecological inference.

## Descriptive Analysis

Now that we've cleaned and prepared the data, we're ready to begin analyzing it. Before running ecological inference, it's useful to conduct some simple descriptive analyses. First, let's plot the precinct-level bivariate relationships between all the different combinations of racial and candidate turnout proportions.

```{r fig.height = 5, fig.width = 7, fig.align = "center"}
plot_bivariate(
  data = gwinnett_ei,
  cand_cols = cands,
  race_cols = races
)
```


The `plot_bivariate()` function returns a `ggplot2` object plotting the relationships between each candidate's precinct-level vote share and the precinct-level turnout share of each race group. These plots can tell us a lot about what to expect from the EI analysese to come. In particular, we can observe the following two trends.

1. We observe what looks like racially polarized voting in the top left corner of this figure. The panel in the upper left corner shows that as the proportion of white voters in a precinct rises, so does the proportion of votes for Brian Kemp. The plot just below shows that the opposite is true for Stacey Abrams. Meanwhile, the second column reveals the opposite pattern for black voters. This is a sign that racially polarized voting might be going on in this election. We can use EI to back this up statistically.

2. The bottom row shows that Jim Metz received very few votes across precincts. Because there is so little variation in levels of vote share for this candidate, our EI estimates probably will not detect racially polarized voting for this candidate. We can expect similarly from the 'hispanic' and 'other' race categories, since few districts have very high proportions of these groups.

These basic descriptive checks help us understand our data and give us a sense of what to expect in our final results. Next, we turn to executing and comparing the different EI techniques.

## Ecological Inference

`eiCompare` enables users to analyze and compare the two dominant EI methods, Iterative EI and RxC EI. For more information about these techniques, see the [technical details](technical.details.com) page of our website.

We can run both methods with their respective functions. They take some time to run because they both compute point estimates by sampling from a distribution. To keep it short for this tutorial, we'll randomly sample 40 precincts (i.e. 40 rows) from the dataset and use these as the basis for our estimates. Obviously, in a real-life setting we would run the analyses on the entire dataset.

```{r}
gwinnett_ei_sample <- gwinnett_ei %>%
  sample_n(100)
```


First, we conduct iterative EI using `ei_iter()`.

```{r}
ei_results_iter <- ei_iter(
  data = gwinnett_ei_sample,
  cand_cols = cands,
  race_cols = races,
  totals_col = total,
  name = "Iter",
  samples = 25
)
summary(ei_results_iter)
```

Notice that the `name` field to names our results object. The name entered here is what differentiates results objects from each other when they are summarized or plotted together.  Next, we conduct RxC EI using `ei_rxc()`. 
```{r}
ei_results_rxc <- ei_rxc(
  data = gwinnett_ei_sample,
  cand_cols = cands,
  race_cols = races,
  totals_col = total,
  name = "RxC"
)
summary(ei_results_iter, ei_results_rxc)
```

Finally, we can plot the results of the two methods using plot_eic. This function accepts a list of `eiCompare` objects outputted by `ei_iter()` and `ei_rxc`. It plots the point estimates and 95% credible intervals for each candidate-race pair, for however many objects are passed in. 
```{r}
results <- list(
  "Iterative" = ei_results_iter,
  "RxC" = ei_results_rxc
)

plot_eic(results)
```

Plotting the results of both methods, we can see that clear evidence of racially polarized voting exists in Gwinnett county. We see that both methods show white voters prefer Brian Kemp, while voters in all other race/ethnicity groups prefer Stacey Abrams.
