---
title: 'Geocoding: Voter Addresses'
author: "null"
date: "8/4/2020"
output:
  pdf_document: default
  html_document: default
---

In this vignette, we will walk-through how to geocode a dataset that includes addresses in order to estimate race using BISG.

## What is Geocoding?

One of the first steps to performing ecological inference using eiCompare is geocoding your voter file addresses in order to perform Bayesian Improved Surname Geocoding (BISG).  Geocoding is the process of using an address or place of location to find geocographic coordinates (i.e. latitude/longitude) of that location on a map. In relation to performing BISG, the values of the geographic coordinates are compared to other census data containing self-reported race and ethnicity to determine the likelihood of an individual living in an ecological unit area (i.e. county, block, tract) being of a certain race given their address. This probability is then used to update a prior probability in the BISG analysis. For more information on BISG, please refer to the BISG vignette. 

Below are some steps to help you walk through the process of performing geocoding on your voter file. 

### Step 1: Load R libraries/packages
Each library/package loaded allows you to use certain functions needed to prep your data for geocoding and run the geocoding tool(s).
```{r}
# Load libraries
suppressMessages(c(
  library(stringr),
  library(readr),
  library(tidyverse),
  library(foreach),
  library(parallel),
  library(doParallel),
  library(data.table),
  library(gmodels),
  library(plyr),
  library(censusxy),
  library(tidycensus),
  library(sf)
))
```

### Step 2: Load your voter data.
We are using the Georgia and Fulton county voter registration dataset and geocoding all voter addresses.

```{r}
# Load full dataset for all registered voters in Georgia
path <- "~/shared/BISG_datasets/"

ga_5k <- read_csv(paste(path, "ga_5k_voter_file.csv", sep = ""))
```

### Load Functions
```{r}
source("~/github/geocodeCompare/geocoder_format.R")
source("~/github/eiCompare/R/run_geocoder.R")
```

### Loading Voter Files
*First, check the full GA voter registration file, `ga_full`, to make sure the dataset has properly downloaded.*
*We will be working with a subset 5000 observations for the Gwinett and Fulton County data.*
```{r}
# Check dimensions of the ga_full dataset
dim(ga_5k)
```

```{r}
# Check first 6 rows
head(ga_5k, 6)
```

```{r}
# Check the column names of the file
names(ga_5k)
```

```{r}
# Change column name for the `name` column to `county_name`
names(ga_5k)[colnames(ga_5k) == "name"] <- "county_name"
names(ga_5k)
```

*All voters should be from Gwinnett and Fulton counties or have county_codes 60 and 67.*
```{r}
# Check that the data only has Gwinnett and Fulton county data
# Frequency count
CrossTable(ga_5k$county_code, ga_5k$county_name, digits = 0)
```

```{r}
# Get the dimensions of the dataset
dim(ga_5k)
```
*There are 1,411,782 million registered voters in Gwinnett and Fulton counties.*

```{r}
# Recode names for location varibles to easily use for geocoding.
names(ga_5k)[names(ga_5k) == "residence_house_number"] <- "street_number"
names(ga_5k)[names(ga_5k) == "residence_street_name"] <- "street_name"
names(ga_5k)[names(ga_5k) == "residence_city"] <- "city"
names(ga_5k)[names(ga_5k) == "residence_zipcode"] <- "zipcode"
names(ga_5k)[names(ga_5k) == "residence_street_suffix"] <- "street_suffix"
```

*Check each address variable for voter's residence to clean data if needed.*
```{r}
# Get the number of missing street numbers
length(which(is.na(ga_5k$street_number)))
```
*There are 5 voters that are missing a street number for their address. We will now put these voters with missing street numbers in a separate dataframe.*

```{r}
# Select the addresses without a street number
street_num_miss <- ga_5k[which(is.na(ga_5k$street_number)), ]
```

```{r}
# Check the frequency of specific addresses
table(street_num_miss$street_name)
```

If you are able to readily locate addresses that have typos or errors, great! However, let's take the most frequent address value to understand why many of these addresses are missing. 
```{r}
# Check the frequency of specific addresses
tail(names(sort(table(street_num_miss$street_name))), 1)
```
It looks like 2 addresses have street names that have been labeled, "MISSING ADDRESS". This may be a data entry error from the voter registration office. We will use the geocoder, OpenStreetMaps to relocate these addresses in the `street_num_miss` dataframe since it takes in addresses later in the tutorial.

Next, we will look at patterns in city, state, and zipcode.
```{r}
# Check the frequency of city names
table(sort(ga_5k$city))
```
```{r}
# Check the frequency of zipcodes
head(table(sort(ga_5k$zipcode)), 100)
```

```{r}
length(which(is.na(ga_5k$zipcode) == TRUE))
```
11 voters are missing zipcodes.

```{r}
length(which(is.na(ga_5k$zipcode) == TRUE & ga_5k$street_name == "MISSING ADDRESS"))
```
6 voters are missing zipcodes and street names.

```{r}
zip_miss <- which(is.na(ga_5k$zipcode) == TRUE & ga_5k$street_name == "MISSING ADDRESS")
ga_5k <- ga_5k[-zip_miss, ]
nrow(ga_5k)
```
There are now 4,989 registered voters in the voter file.

There are also 9-digit zipcodes without hyphens in the voter file. Having zipcodes as hyphens may help the goeocoder be more accurate when trying to find an individual's corrdinates.
```{r}
source("~/github/geocodeCompare/geocoder_format.R")
# Format zipcodes that are 9 digits long and treated as character variables
ga_5k <- zip_hyphen(
  voter_file = ga_5k,
  voter_id = "registration_number",
  zipcode = "zipcode"
)
```


```{r}
# Check that zzipcodes have been properly formatted
head(table(ga_5k$zipcode), 6)
```

### Step 3: Prepare/Structure your voter data for geocoding.

*Concatenate columns for street address.*
```{r}
ga_5k <- concat_streetname(
  voter_file = ga_5k,
  voter_id = "registration_number",
  street_number = "street_number",
  street_suffix = "street_suffix"
)

head(ga_5k$street_address, 6)
```

*Create a column for the final address in the voter file.*
```{r}
# Create a column for state. The GA voter registration dataset does not have a state column.
ga_5k$state <- "GA"

ga_5k <- concat_final_address(
  voter_file = ga_5k,
  voter_id = "registration_number",
  street_address = "street_address",
  city = "city",
  state = "state",
  zipcode = "zipcode"
)
```

```{r}
# convert dataframe into a tibble
ga_5k <- as_tibble(ga_5k)

head(ga_5k, 6)
```


### Step 4: run_geocoder() Function 
### Select a geocoder and run the geocoder on the addresses in your file.

*Select the geocoder you are going to use to find the geographies like coordinates (i.e. latitude and longitude) and FIPS codes for the addresses in the voter file.* 

#### There are several options for geocoding your data using a geocoding API.  The eiCompare package utilizes the US Census Geocding API via a R package called censusxy.


*Note: If you have more than 10000 voters in yoru file, we recommedn using parallel processing. More information on parallel processing can be found on the Parallel Processing vignette.*

#### Let's start geocoding our data*
*We recommend first geocoding your data with the US Census Geocoder API via the R package, censusxy.*

voter file: A data frame contain the voter addresses, separated into columns for street, city, state, and zipcode

geocoder: The options for selecting geocoders are "censusxy" and "opencage".

parallel: TRUE or FALSE. The option to run parallel processing on the data. This is dependent on whether your dataset exceeds 10,000 observations (i.e. addresses). Geocoding a batch of 10000 addresses is also the limit for addresses to be geocoded with the US Census Geocoder API.

voter_id: the unique identifier 

street: the street number, street name, and/or street suffix. Ex. 555 Main Street SW

city: the location/town

state: the abbreviated state (U.S. state categories such as "GA")

zipcode: the 5 or 9 digit number in the format XXXXX or XXXXX-XXXX.

country: the abbreviated a nation or territory

census_return: either "locations" or "geographies". "locations" returns the latitude and longitude coordinates. "geographies" returns the latitude, longitude, and FIPS codes for county, state, tract, and block.

census_benchmark: a dataset of the snapshot of the US Census data. Data is collected two times a year. Public_AR_Current is the time period when we created the snapshot of the data (usually done twice yearly). For example, Public_AR_Current is the most recent snapshot of our dataset. Other options can be found at: _____<insert link>____________________.

census_vintage: a dataset based on the census


We only have 4995 observations. So, we will not run parallel processing and set parallel to FALSE in the run_geocoder() function.



*If there are any missing geocoded addresses, the run_geocoder() function will re-run the geocoder on those missing geocoded addresses. Then, join the missing data back in with the original voter file.*

```{r}
geocoded_data <- run_geocoder(
  voter_file = ga_5k,
  geocoder = "census",
  parallel = FALSE,
  voter_id = "registration_number",
  street = "street_address",
  city = "city",
  state = "state",
  zipcode = "zipcode",
  country = "US",
  census_return = "geographies",
  census_benchmark = "Public_AR_Current",
  census_output = "full",
  census_class = "sf",
  census_vintage = 4,
  opencage_key = NULL
)
```

Check the column names of the geocoded dataset. There should be an additional column called `geometry` with latitude, longitude coordinates.
```{r}

# Check the column names of the geocoded_dataset object
names(geocoded_data)
```

```{r}
# Check the first six rows of the geocoded_dataset object
head(geocoded_data)
```

```{r}
# The number of rows missing in new dataframe
num_miss_geo <- nrow(ga_5k) - nrow(geocoded_data)

# Only re-run the geocoder if missing data is present.
if (num_miss_geo > 0) {

  # Find non-geocoded data
  missing_lonlat_df <- anti_join(ga_5k, as.data.frame(geocoded_data))

  # Run the geocoder on the missing data
  rerun_data <- run_geocoder(
    voter_file = missing_lonlat_df,
    geocoder = "census",
    parallel = FALSE,
    voter_id = "registration_number",
    street = "street_address",
    city = "city",
    state = "state",
    zipcode = "zipcode",
    country = "US",
    census_return = "geographies",
    census_benchmark = "Public_AR_Current",
    census_output = "full",
    census_vintage = 4,
    census_class = "sf",
    opencage_key = NULL
  )
}
```

*Combine the newly geocoded data from the `rerun_data` object and the original geocoded data, `geocoded_data`, object.*
```{r}
geo_combined <- rbind(geocoded_data, rerun_data)
```

```{r}
# Check the number of observations in the combined geocoded voter registration dataset
nrow(geo_combined)
```

```{r}
# Check the dimensions of the combined geocoded voter registration dataset
dim(geo_combined)
```
```{r}
# rename columns
names(geo_combined)[names(geo_combined) == "cxy_state_id"] <- "STATEFP10"
names(geo_combined)[names(geo_combined) == "cxy_county_id"] <- "COUNTYFP10"
names(geo_combined)[names(geo_combined) == "cxy_tract_id"] <- "TRACTCE10"
names(geo_combined)[names(geo_combined) == "cxy_block_id"] <- "BLOCKCE10"
```

### Step 6: Save your geocoded file 
```{r}
write_csv(geo_combined, "~/shared/BISG_datasets/gwin_fulton_5k_fullgeo.csv")
```

### Step 7: Plot your geocoded data
```{r}
# interactive map
library(mapview)
mapview::mapview(geo_combined)
```

### Step 8: Mapping more data
```{r}
# J to Insert interactive map function here.
```
      
