---
title: 'Geocoding: Voter Addresses'
author: "null"
date: "8/4/2020"
output:
  pdf_document: default
  html_document: default
---

In this vignette, we will walk-through how to geocode a dataset that includes addresses in order to estimate race using BISG.

## What is Geocoding?

One of the first steps to performing ecological inference using eiCompare is geocoding your voter file addresses in order to perform Bayesian Improved Surname Geocoding (BISG).  Geocoding is the process of using an address or place of location to find geocographic coordinates (i.e. latitude/longitude) of that location on a map. In relation to performing BISG, the values of the geographic coordinates are compared to other census data containing self-reported race and ethnicity to determine the likelihood of an individual living in an ecological unit area (i.e. county, block, tract) being of a certain race given their address. This probability is then used to update a prior probability in the BISG analysis. For more information on BISG, please refer to the BISG vignette. 

Below are some steps to help you walk through the process of performing geocoding on your voter file. 

### Step 1: Load R libraries/packages
Each library/package loaded allows you to use certain functions needed to prep your data for geocoding and run the geocoding tool(s).
```{r}
# Load libraries
suppressMessages(c(
  library(stringr),
  library(readr),
  library(tidyverse),
  library(foreach),
  library(parallel),
  library(doParallel),
  library(data.table),
  library(gmodels),
  library(plyr),
  library(censusxy),
  library(tidycensus),
  library(mapview),
  library(tigris),
  library(sf)
))
```

### Step 2: Load your voter data.
We are using the Georgia and Fulton county voter registration dataset and geocoding all voter addresses.

```{r}
# Load full dataset for all registered voters in Georgia
path <- "~/shared/BISG_datasets/"

ga_5k <- read_csv(paste(path, "ga_5k_voter_file.csv", sep = ""))
```

### Load Functions
```{r}
source("~/github/geocodeCompare/geocoder_format.R")
source("~/github/eiCompare/R/run_geocoder.R")
source("~/github/eiCompare/R/map_shapefile.R")
```

### Loading Voter Files
*First, check the full GA voter registration file, `ga_full`, to make sure the dataset has properly downloaded.*
*We will be working with a subset 5000 observations for the Gwinett and Fulton County data.*
```{r}
# Check dimensions of the ga_full dataset
dim(ga_5k)
```

```{r}
# Check first 6 rows
head(ga_5k, 6)
```

```{r}
# Check the column names of the file
names(ga_5k)
```

```{r}
# Change column name for the `name` column to `county_name`
names(ga_5k)[colnames(ga_5k) == "name"] <- "county_name"
names(ga_5k)
```

*All voters should be from Gwinnett and Fulton counties or have county_codes 60 and 67.*
```{r}
# Check that the data only has Gwinnett and Fulton county data
# Frequency count
CrossTable(ga_5k$county_code, ga_5k$county_name, digits = 0)
```

```{r}
# Get the dimensions of the dataset
dim(ga_5k)
```
*There are 1,411,782 million registered voters in Gwinnett and Fulton counties.*

```{r}
# Recode names for location varibles to easily use for geocoding.
names(ga_5k)[names(ga_5k) == "residence_house_number"] <- "street_number"
names(ga_5k)[names(ga_5k) == "residence_street_name"] <- "street_name"
names(ga_5k)[names(ga_5k) == "residence_city"] <- "city"
names(ga_5k)[names(ga_5k) == "residence_zipcode"] <- "zipcode"
names(ga_5k)[names(ga_5k) == "residence_street_suffix"] <- "street_suffix"
```

*Check each address variable for voter's residence to clean data if needed.*
```{r}
# Get the number of missing street numbers
length(which(is.na(ga_5k$street_number)))
```
*There are 5 voters that are missing a street number for their address. We will now put these voters with missing street numbers in a separate dataframe.*

```{r}
# Select the addresses without a street number
street_num_miss <- ga_5k[which(is.na(ga_5k$street_number)), ]
```

```{r}
# Check the frequency of specific addresses
table(street_num_miss$street_name)
```

If you are able to readily locate addresses that have typos or errors, great! However, let's take the most frequent address value to understand why many of these addresses are missing. 
```{r}
# Check the frequency of specific addresses
tail(names(sort(table(street_num_miss$street_name))), 1)
```
It looks like 2 addresses have street names that have been labeled, "MISSING ADDRESS". This may be a data entry error from the voter registration office. We will use the geocoder, OpenStreetMaps to relocate these addresses in the `street_num_miss` dataframe since it takes in addresses later in the tutorial.

Next, we will look at patterns in city, state, and zipcode.
```{r}
# Check the frequency of city names
table(sort(ga_5k$city))
```
```{r}
# Check the frequency of zipcodes
head(table(sort(ga_5k$zipcode)), 100)
```

```{r}
length(which(is.na(ga_5k$zipcode) == TRUE))
```
11 voters are missing zipcodes.

```{r}
length(which(is.na(ga_5k$zipcode) == TRUE & ga_5k$street_name == "MISSING ADDRESS"))
```
6 voters are missing zipcodes and street names.

```{r}
zip_miss <- which(is.na(ga_5k$zipcode) == TRUE & ga_5k$street_name == "MISSING ADDRESS")
ga_5k <- ga_5k[-zip_miss, ]
nrow(ga_5k)
```
There are now 4,989 registered voters in the voter file.

There are also 9-digit zipcodes without hyphens in the voter file. Having zipcodes as hyphens may help the goeocoder be more accurate when trying to find an individual's corrdinates.
```{r}
source("~/github/geocodeCompare/geocoder_format.R")
# Format zipcodes that are 9 digits long and treated as character variables
ga_5k <- zip_hyphen(
  voter_file = ga_5k,
  voter_id = "registration_number",
  zipcode = "zipcode"
)
```


```{r}
# Check that zzipcodes have been properly formatted
head(table(ga_5k$zipcode), 6)
```

### Step 3: Prepare/Structure your voter data for geocoding.

*Concatenate columns for street address.*
```{r}
ga_5k <- concat_streetname(
  voter_file = ga_5k,
  voter_id = "registration_number",
  street_number = "street_number",
  street_suffix = "street_suffix"
)

head(ga_5k$street_address, 6)
```

*Create a column for the final address in the voter file.*
```{r}
# Create a column for state. The GA voter registration dataset does not have a state column.
ga_5k$state <- "GA"

ga_5k <- concat_final_address(
  voter_file = ga_5k,
  voter_id = "registration_number",
  street_address = "street_address",
  city = "city",
  state = "state",
  zipcode = "zipcode"
)
```

```{r}
# convert dataframe into a tibble
ga_5k <- as_tibble(ga_5k)

head(ga_5k, 6)
```


### <u>Step 4: run_geocoder() Function<u>

### Select a geocoder and run the geocoder on the addresses in your file.

Select the geocoder you are going to use to find the geographies like coordinates (i.e. latitude and longitude) and FIPS codes for the addresses in the voter file. There are several options for geocoding your data using a geocoding API.  The eiCompare package utilizes the US Census Geocding API via a R package called censusxy. For an alternative commerically available geocoder, we recommend using Opencage Geocoder API which has limits of 2500 requests per day.

*Note: If you have more than 10000 voters in your file, we recommend using parallel processing. More information on parallel processing can be found on the Parallel Processing vignette.*

#### Let's start geocoding our data
*We recommend first geocoding your data with the US Census Geocoder API via the R package, censusxy.*

The US Census Geocoder has two options for geocoding output: "simple" and "full".
+ "simple" returns coordinates of latitude (lat) and longitude. 
+ "full" returns coordinates, and other variables for geographies from Federal Information Processing Standards (FIPS) codes.

To get the latitude and longitude only, we will set the `census_return` variable to "simple" and assign each census variable to a desired value. We only have 4995 observations. So, we will not run parallel processing and set parallel to FALSE in the run_geocoder() function.

```{r}
# Getting the latitude and longitude coordinates only.
geocoded_data_simple <- run_geocoder(
  voter_file = ga_5k,
  geocoder = "census",
  parallel = FALSE,
  voter_id = "registration_number",
  street = "street_address",
  city = "city",
  state = "state",
  zipcode = "zipcode",
  country = "US",
  census_return = "geographies",
  census_benchmark = "Public_AR_Current",
  census_output = "simple",
  census_class = "sf",
  census_vintage = 4,
  opencage_key = NULL
)
```
Check the column names of the geocoded dataset. There should be an additional column called `geometry` with latitude, and longitude coordinates.

```{r}
colnames(geocoded_data_simple)
```

Next, we will use parallel processing to make our geocoder run faster by setting `parallel`=TRUE and obtain simple geographies.
```{r}
# Getting the latitude and longitude coordinates only.
geocoded_data_simple <- run_geocoder(
  voter_file = ga_5k,
  geocoder = "census",
  parallel = TRUE,
  voter_id = "registration_number",
  street = "street_address",
  city = "city",
  state = "state",
  zipcode = "zipcode",
  country = "US",
  census_return = "geographies",
  census_benchmark = "Public_AR_Current",
  census_output = "simple",
  census_class = "sf",
  census_vintage = 4,
  opencage_key = NULL
)
```

We will now demonstarte how to add the the coordinates and FIPS codes by setting the `census_return` variable to "full".
```{r}
geocoded_data_full_geo <- run_geocoder(
  voter_file = ga_5k,
  geocoder = "census",
  parallel = FALSE,
  voter_id = "registration_number",
  street = "street_address",
  city = "city",
  state = "state",
  zipcode = "zipcode",
  country = "US",
  census_return = "geographies",
  census_benchmark = "Public_AR_Current",
  census_output = "full",
  census_class = "sf",
  census_vintage = 4,
  opencage_key = NULL
)
```

Check the column names of the geocoded dataset. There should be an additional column called `geometry` with latitude, longitude coordinates, and other variables for geographies.
```{r}

# Check the column names of the geocoded_dataset object
colnames(geocoded_data_full_geo)
```

```{r}
# Check the first six rows of the geocoded_dataset object
head(geocoded_data_full_geo)
```


If there are any missing geocoded addresses, use the run_geocoder() function to re-run the geocoder on those missing geocoded addresses. We will use the `geocoded_data_full_geo` data to demonstrate how to re-run the geocoder on missing addresses.
```{r}
# The number of rows missing in new dataframe
num_miss_geo <- nrow(ga_5k) - nrow(geocoded_data_full_geo)

# Only re-run the geocoder if missing data is present.
if (num_miss_geo > 0) {

  # Find non-geocoded data
  missing_lonlat_df <- anti_join(ga_5k, as.data.frame(geocoded_data))

  # Run the geocoder on the missing data
  rerun_data <- run_geocoder(
    voter_file = missing_lonlat_df,
    geocoder = "census",
    parallel = FALSE,
    voter_id = "registration_number",
    street = "street_address",
    city = "city",
    state = "state",
    zipcode = "zipcode",
    country = "US",
    census_return = "geographies",
    census_benchmark = "Public_AR_Current",
    census_output = "full",
    census_vintage = 4,
    census_class = "sf",
    opencage_key = NULL
  )
}
```

Some of the missing addresses were able to be geocoded. Next, we will combine the newly geocoded data from the `rerun_data` object and the original geocoded data, `geocoded_data`, object.
```{r}
geo_combined <- rbind(geocoded_data_full_geo, rerun_data)
```

```{r}
# Check the number of observations in the combined geocoded voter registration dataset
nrow(geo_combined)
```

```{r}
# Check the dimensions of the combined geocoded voter registration dataset
dim(geo_combined)
```

```{r}
# rename columns to US Census FIPS code variable names
names(geo_combined)[names(geo_combined) == "cxy_state_id"] <- "STATEFP10"
names(geo_combined)[names(geo_combined) == "cxy_county_id"] <- "COUNTYFP10"
names(geo_combined)[names(geo_combined) == "cxy_tract_id"] <- "TRACTCE10"
names(geo_combined)[names(geo_combined) == "cxy_block_id"] <- "BLOCKCE10"
```

### Step 6: Save your geocoded file 
```{r}
write_csv(geo_combined, "~/shared/BISG_datasets/gwin_fulton_5k_fullgeo.csv")
```

### Step 7: Plot your geocoded data
We will map the area or ecological unit we are interested in using the tigris package for loading in US Census shapefiles.
```{r}
# Load shapefile for the state of Georgia
shape_file <- counties(state = "GA")

# Concatenate the state and county codes into column called fips
shape_file$fips <- paste0(shape_file$STATEFP, shape_file$COUNTYFP)

# Filter shape_file for the counties: Gwinnett and Fulton
shape_file <- shape_file[shape_file$fips == "13121" | shape_file$fips == "13135", ]
shape_file$fulton <- ifelse(shape_file$fips == "13121", 1, 0)
shape_file$gwinnett <- ifelse(shape_file$fips == "13121", 1, 0)

# Map shape_file
county_shape <- map_shape_file(
  shape_file = shape_file,
  crs = "+proj=latlong +ellps=GRS80 +no_defs",
  title = "Gwinnett and Fulton counties"
)
county_shape
```
We now will look at the block level of Fulton and Gwinnett county
```{r}
# Load shape file using tidycensus
gwin_fulton_blocks <- blocks(state = "GA", county = c("Gwinnett", "Fulton"))
# Concatenate the state and county codes into column called fips
gwin_fulton_blocks$fips <- paste0(gwin_fulton_blocks$STATEFP, gwin_fulton_blocks$COUNTYFP)

gwin_fulton_map <- map_shape_points(
  voter_file = geo_combined,
  shape_file = gwin_fulton_blocks,
  crs = "+proj=longlat +ellps=GRS80",
  title = "Gwinnett and Fulton Counties - All Registered Voters"
)
gwin_fulton_map
```


### Step 8: Mapping more data
```{r}
# Assign the sf object to a variable name inter_geo
inter_geo <- geocoded_data_full_geo

# Create interactive map using mapview
mapview::mapview(inter_geo)
```
      
