---
title: "Geocoding: Voter Addresses"
author: 
date: "8/4/2020"
output: html_document
---

In this vignette, we will walk-through how to geocode a dataset that includes addresses in order to estimate race using BISG.

## What is Geocoding?

One of the first steps to performing ecological inference using eiCompare is geocoding your voter file addresses in order to perform Bayesian Improved Surname Geocoding (BISG).  Geocoding is the process of using an address or place of location to find geocographic coordinates (i.e. latitude/longitude) of that location on a map. In relation to performing BISG, the values of the geographic coordinates are compared to other census data containing self-reported race and ethnicity to determine the likelihood of an individual living in an ecological unit area (i.e. county, block, tract) being of a certain race given their address. This probability is then used to update a prior probability in the BISG analysis. For more information on BISG, please refer to the BISG vignette. 

Below are some steps to help you walk through the process of performing geocoding on your voter file. 

### Step 1: Load R libraries/packages
Each library/package loaded allows you to use certain functions needed to prep your data for geocoding and run the geocoding tool(s).
```{r}
# Load libraries
suppressMessages(
  c(
    install.packages("stringr"),
    library(stringr), # US Census geocoding API package for R
    install.packages("readr"),
    library(readr), # for loading csv file
    install.packages("tidyverse"),
    library(tidyverse), # for making a tibble from a dataframe,
    install.packages("foreach"),
    library(foreach),
    install.packages("parallel"),
    library(parallel),
    install.packages("doParallel"),
    library(doParallel),
    install.packages("data.table"),
    library(data.table),
    install.packages("gmodels"),
    library(gmodels)
  )
)
```

### Step 2: Load your voter data.
We are using the Georgia and Fulton county voter registration dataset and geocoding all voter addresses.

```{r}
# Load full dataset for all registered voters in Georgia
ga_full <- read_csv("~/shared/BISG_datasets/georgia_full_voterfile.csv")

# Load county code data
ga_county_codes <- read_csv("~/shared/BISG_datasets/georgia-counties.csv")
```

*First, check the full GA voter registration file, `ga_full`, to make sure the dataset has properly downloaded.*
```{r}
# Check first 6 rows
head(ga_full, 6)
```

```{r}
# Check the column names of the file
names(ga_full)
```

```{r}
# Check dimensions of the ga_full dataset
dim(ga_full)
```


*Now, check the Georgia county code reference table to make sure the file has been properly loaded.*
```{r}
# Check first 6 rows
head(ga_county_codes, 6)
```

```{r}
# Check the column names of the file
names(ga_county_codes)
```


*There should be 159 counties in the state of Georgia.*
```{r}
# Check dimensions of the county code dataset
dim(ga_county_codes)
```
### *Merge the `ga_full` dataset with county names in the `ga_county_codes` table*
```{r}
# Merge `ga_full` and `ga_county_codes`
ga_full_counties <- merge(ga_full, ga_county_codes, by = "county_code")
```

*Next, we will subset the `ga_full` voter registration dataframe to only include information on people from Gwinnett and Fulton counties, using the column name `county_code`. `county code` is an array that assigns a number to each county name. For Fulton County, the county_code is 60 and Gwinnett County, the county_code is 67.*

*Note: This code is different than the FIPS code used to identify counties within census data.*

```{r}
ga_gwin_fulton <- ga_full_counties[which(ga_full_counties$county_code == 60 | ga_full_counties$county_code == 67), ]

# Check the first 6 rows of the dataset
head(ga_gwin_fulton)
```

```{r}
# Obtain column names for the ga_gwin_fulton dataset
names(ga_gwin_fulton)
```

```{r}
# Change column name for the `name` column to `county_name`
names(ga_gwin_fulton)[colnames(ga_gwin_fulton) == "name"] <- "county_name"
names(ga_gwin_fulton)
```

*All voters should be from Gwinnett and Fulton counties or have county_codes 60 and 67.*
```{r}
# Check that the data only has Gwinnett and Fulton county data
# Frequency count
CrossTable(ga_gwin_fulton$county_code, ga_gwin_fulton$county_name, digits = 0)
```

```{r}
# Get the dimensions of the dataset
dim(ga_gwin_fulton)
```
*There are 1,411,782 million registered voters in Gwinnett and Fulton counties.*

```{r}
# Recode names for location varibles to easily use for geocoding.
names(ga_gwin_fulton)[names(ga_gwin_fulton) == "residence_house_number"] <- "street_number"
names(ga_gwin_fulton)[names(ga_gwin_fulton) == "residence_street_name"] <- "street_name"
names(ga_gwin_fulton)[names(ga_gwin_fulton) == "residence_city"] <- "city"
names(ga_gwin_fulton)[names(ga_gwin_fulton) == "residence_zipcode"] <- "zipcode"
names(ga_gwin_fulton)[names(ga_gwin_fulton) == "residence_street_suffix"] <- "street_suffix"
```

*Check each address variable for voter's residence to clean data if needed.*
```{r}
# Get the number of missing street numbers
length(which(is.na(ga_gwin_fulton$street_number)))
```
*There are 625 voters that are missing a street number for their address. We will now put these voters with missing street numbers in a separate dataframe.*

```{r}
# Select the addresses without a street number
street_num_miss <- ga_gwin_fulton[which(is.na(ga_gwin_fulton$street_number)), ]
```

```{r}
# Check the frequency of specific addresses
table(street_num_miss$street_name)
```

If you are able to readily locate addresses that have typos or errors, great! However, let's take the most frequent address value to understand why many of these addresses are missing. 
```{r}
# Check the frequency of specific addresses
tail(names(sort(table(street_num_miss$street_name))), 1)
```
It looks like 304 addresses have street names that have been labeled, "MISSING ADDRESS". This may be a data entry error from the voter registration office. We will use the geocoder, OpenStreetMaps to relocate these addresses in the `street_num_miss` dataframe since it takes in addresses later in the tutorial.

Next, we will look at patterns in city, state, and zipcode.
```{r}
# Check the frequency of city names
table(sort(ga_gwin_fulton$city))
```
```{r}
# Check the frequency of zipcodes
head(table(sort(ga_gwin_fulton$zipcode)), 100)
```

```{r}
length(which(is.na(ga_gwin_fulton$zipcode) == TRUE))
```
```{r}
length(which(is.na(ga_gwin_fulton$zipcode) == TRUE & ga_gwin_fulton$street_name == "MISSING ADDRESS"))
```

```{r}
zip_miss <- which(is.na(ga_gwin_fulton$zipcode) == TRUE & ga_gwin_fulton$street_name == "MISSING ADDRESS")
ga_gwin_fulton <- ga_gwin_fulton[-zip_miss, ]
nrow(ga_gwin_fulton)
```
There are now 1,409,811 registered voters in the voter file.

There are also 9 digit zipcodes without hyphens in the viter file. Having zipcodes as hyphens may help the goeocoder be more accurate when trying to find an individual's corrdinates.
```{r}
# Format zipcodes that are 9 digits long and treated as character variables
ga_gwin_fulton <- zip_hyphen(
  voter_file = ga_gwin_fulton,
  voter_id = "registration_number",
  zipcode = "zipcode"
)
```


```{r}
# Check that zzipcodes have been properly formatted
head(table(ga_gwin_fulton$zipcode), 6)
```

### Step 3: Prepare/Structure your voter data for geocoding.

*Concatenate columns for street address.*
```{r}
ga_gwin_fulton <- concat_streetname(
  voter_file = ga_gwin_fulton,
  voter_id = "registration_number",
  street_number = "street_number",
  street_suffix = "street_suffix"
)

head(ga_gwin_fulton$street_address, 6)
```

*Create a column for the final address in the voter file.*
```{r}
ga_gwin_fulton <- concat_final_address(
  voter_file = ga_gwin_fulton,
  voter_id = "registration_number",
  street_address = "street_address",
  city = "city",
  state = "GA",
  zipcode = "zipcode"
)
```

*
```{r}
# convert dataframe into a tibble
ga_gwin_fulton <- as_tibble(ga_gwin_fulton)

head(ga_gwin_fulton, 6)
```

### Step 4: Select a geocoder and run the geocoder on the addresses in your file.

*Select the geocoder you are going to use to find the coordinates (i.e. latitude and longitude) for the addresses in the voter file.* 

*Note: If you have more than 10000 voters in yoru file, we recommedn using parallel processing. More information on parallel processing can be found on the Parallel Processing vignette.*

*We recommend first geocoding your data with the US Census Geocoder API via the R package, censusxy.* 

#### Using the Census Geocoder API

*Let's start geocoding our data*


### Step 5: Re-run the geocoder on missing geocoded addresses.

Determine if there are any missing geocoded addresses and re-run the geocoder on those missing geocoded addresses. Then, join the missing data back with the original voter file.


### Step 6: Save your geocoded file 
