---
title: "Racially Polarized Voting Analysis with eiCompare"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Racially Polarized Voting Analysis with eiCompare}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
options(scipen = 999, digits = 4)
```

This vignette demonstrates how to conduct a racially polarized voting (RPV)
analysis using `eiCompare` and `eiExpand`. We walk through the full workflow:
descriptive bivariate plots, three ecological inference methods (Goodman's
regression, iterative EI, and RxC EI), and visualization of results using
`eiExpand::rpv_plot()`.

## Data

We use the `gwinnett_ei` dataset bundled with `eiCompare`. This contains
precinct-level data from the 2018 Gwinnett County, Georgia gubernatorial
election between Republican Brian Kemp, Democrat Stacey Abrams, and
Libertarian Jim Metz.

```{r setup, message = FALSE, warning = FALSE}
library(eiCompare)
library(dplyr)

data("gwinnett_ei")
head(gwinnett_ei)
```

The columns are already expressed as proportions of turnout (candidates) and
proportions of the voting population (race groups). The candidate columns sum
to 1 and the race columns sum to 1 within each precinct.

```{r}
cands <- c("kemp", "abrams", "metz")
races <- c("white", "black", "other")
total <- "turnout"
```

## Step 1: Descriptive analysis with bivariate plots

Before running any EI model, it is useful to examine the precinct-level
relationship between each racial group's share of the electorate and each
candidate's vote share. The `plot_bivariate()` function produces a faceted
scatterplot with OLS trend lines.

```{r bivariate-plot}
plot_bivariate(
  data = gwinnett_ei,
  cand_cols = cands,
  race_cols = races,
  corrs = TRUE
)
```

The bivariate plots reveal clear patterns of racially polarized voting:

- **White voters**: As the proportion of white voters in a precinct increases,
  Kemp's vote share rises and Abrams' falls. The correlation between white
  share and Kemp vote share is strongly positive.
- **Black voters**: The opposite pattern holds. Higher black voter share is
  associated with higher Abrams vote share.
- **Metz**: Received very few votes across all precincts, so we do not expect
  EI to detect meaningful racial differences for this candidate.

We can also inspect the correlation matrix directly:

```{r}
race_cand_cors(
  data = gwinnett_ei,
  cand_cols = cands,
  race_cols = races
)
```

## Step 2: Goodman's ecological regression

Goodman's regression is the simplest ecological inference method. It fits a
linear model for each candidate-race pair and extracts the implied group-level
vote share from the regression coefficients. It is fast and deterministic but
can produce estimates outside the [0, 1] bound.

As of version 3.0.6, `ei_good()` returns an `eiCompare` class object,
consistent with `ei_iter()` and `ei_rxc()`.

```{r goodman}
ei_good_out <- ei_good(
  data = gwinnett_ei,
  cand_cols = cands,
  race_cols = races,
  totals_col = total
)

summary(ei_good_out)
```

The summary shows point estimates and standard errors for each candidate-race
combination. Note that Goodman estimates are based on OLS and do not produce
credible intervals or posterior samples.

## Step 3: Iterative EI (King's 2x2 method)

Iterative EI runs King's (1997) 2x2 ecological inference model for every
candidate-race pair. It produces Bayesian point estimates with credible
intervals.

```{r ei-iter, results = "hide", message = FALSE}
ei_iter_out <- ei_iter(
  data = gwinnett_ei,
  cand_cols = cands,
  race_cols = races,
  totals_col = total,
  name = "Iter",
  seed = 543210
)
```

```{r}
summary(ei_iter_out)
```

## Step 4: RxC EI (multinomial-Dirichlet)

RxC EI estimates all candidate-race combinations simultaneously using a
Bayesian multinomial-Dirichlet model. This method accounts for the constraint
that vote shares must sum to one within each racial group.

```{r ei-rxc, results = "hide", message = FALSE, warning = FALSE}
ei_rxc_out <- ei_rxc(
  data = gwinnett_ei,
  cand_cols = cands,
  race_cols = races,
  totals_col = total,
  name = "RxC",
  seed = 543210,
  samples = 10000,
  burnin = 5000
)
```

```{r}
summary(ei_rxc_out)
```

## Step 5: Compare EI methods

The built-in `plot()` method for `eiCompare` objects lets us visually compare
estimates across methods:

```{r compare-plot, fig.height = 6, fig.width = 8}
plot(ei_iter_out, ei_rxc_out)
```

## Step 6: Prepare results for RPV visualization

The `rpv_toDF()` function converts EI results into a tidy data frame suitable
for `eiExpand::rpv_plot()`. We prepare data frames for all three methods.

```{r rpv-todf, message = FALSE, warning = FALSE}
# Goodman
good_df <- rpv_toDF(
  rpv_results = ei_good_out,
  model = "Goodman",
  jurisdiction = "Gwinnett",
  candidate = c("Kemp", "Abrams", "Metz"),
  preferred_candidate = c("White", "Black", "Other"),
  party = c("Republican", "Democratic", "Libertarian"),
  election_type = "General",
  year = "2018",
  contest = "Governor"
)

# Iterative EI
iter_df <- rpv_toDF(
  rpv_results = ei_iter_out,
  model = "Iter EI",
  jurisdiction = "Gwinnett",
  candidate = c("Kemp", "Abrams", "Metz"),
  preferred_candidate = c("White", "Black", "Other"),
  party = c("Republican", "Democratic", "Libertarian"),
  election_type = "General",
  year = "2018",
  contest = "Governor"
)

# RxC EI
rxc_df <- rpv_toDF(
  rpv_results = ei_rxc_out,
  model = "RxC EI",
  jurisdiction = "Gwinnett",
  candidate = c("Kemp", "Abrams", "Metz"),
  preferred_candidate = c("White", "Black", "Other"),
  party = c("Republican", "Democratic", "Libertarian"),
  election_type = "General",
  year = "2018",
  contest = "Governor"
)
```

We can inspect the structure:

```{r}
head(iter_df)
```

## Step 7: Visualize RPV results with eiExpand

The `eiExpand` package provides `rpv_plot()` for publication-quality RPV bar
charts. We can plot each method individually or combine them.

```{r rpv-plot-iter, message = FALSE, warning = FALSE, fig.height = 5, fig.width = 8}
library(eiExpand)

rpv_plot(iter_df,
  title = "Racially Polarized Voting: Iterative EI",
  subtitle = "2018 Gwinnett County Governor"
)
```

```{r rpv-plot-rxc, message = FALSE, warning = FALSE, fig.height = 5, fig.width = 8}
rpv_plot(rxc_df,
  title = "Racially Polarized Voting: RxC EI",
  subtitle = "2018 Gwinnett County Governor"
)
```

We can also combine all three methods into one data frame for a comparative
plot:

```{r rpv-plot-combined, message = FALSE, warning = FALSE, fig.height = 6, fig.width = 9}
combined_df <- dplyr::bind_rows(good_df, iter_df, rxc_df)

rpv_plot(combined_df,
  title = "RPV Estimates Across Methods",
  subtitle = "2018 Gwinnett County Governor",
  panelBy = "Model"
)
```

## Step 8: VAP-denominator normalization with rpv_normalize

When using voting age population (VAP) or citizen voting age population
(CVAP) as the denominator instead of total votes cast, we include a "no vote"
category in the candidate columns. After EI estimation, `rpv_normalize()`
conditions the estimates on people who actually voted, removing the no-vote
share.

This approach is useful when precinct-level voter file data is unavailable
and demographic denominators (VAP/CVAP) are used instead.

**Note:** `rpv_normalize()` requires posterior samples from MCMC-based methods
and therefore works with `ei_iter()` and `ei_rxc()` output only. Passing
`ei_good()` output will produce an error, since Goodman's regression does not
generate posterior samples.

```{r rpv-normalize-error}
# This will error as expected:
tryCatch(
  rpv_normalize(
    ei_object = ei_good_out,
    cand_cols = c("kemp", "abrams"),
    race_cols = c("white", "black")
  ),
  error = function(e) message("Expected error: ", e$message)
)
```

## Summary

This vignette demonstrated the full RPV analysis workflow:

1. **Descriptive analysis** with `plot_bivariate()` and `race_cand_cors()`
2. **Goodman's regression** with `ei_good()` -- fast baseline estimates
3. **Iterative EI** with `ei_iter()` -- Bayesian 2x2 estimates with CIs
4. **RxC EI** with `ei_rxc()` -- simultaneous multinomial-Dirichlet estimates
5. **Method comparison** with `plot()` and side-by-side `rpv_toDF()` output
6. **RPV visualization** with `eiExpand::rpv_plot()`

All three EI functions return `eiCompare` class objects that work with
`summary()`, `rpv_toDF()`, and `plot()`. The `rpv_toDF()` output is directly
compatible with `eiExpand::rpv_plot()` for publication-ready RPV bar charts.
